A seamless, client-centric programming model for type safe web applications

### 2014-05-07 paper.tex
Useless -> Remote, main is entry point

\begin{itemize}
  \item The programming model is synchronous, giving the programmer a simple,
        linear view of the program flow, eliminating the need to program with
        callbacks and continuations.
  \item Side-effecting code is explicitly designated to run on either the
        client or the server using the type system while pure code can be
        shared by both. Additionally, general IO computations may be lifted
        into both client and server code, allowing for safe IO code reuse
        within the confines of the client or server designated functions.
  \item Client-server network communication is handled through statically typed
        RPC function calls, extending the reach of Haskell's type checker over
        the network and giving the programmer advance warning when she uses
        network services incorrectly or forgets to update communication code
        as the application's internal protocol changes.
  \item Our model takes the view that the client side is the main driver when
        developing web applications and accordingly delegates the server to
        the role of a computational and/or storage resource, tasked with
        servicing client requests rather than driving the program. While it is
        entirely possible to implement a server-to-client communication channel
        on top of our model, we believe that choosing one side of the
        heterogenous client-server relation as the master helps keeping the
        program flow linear and predictable.
  \item The implementation is built as a library on top of the GHC and Haste
        Haskell compilers, requiring little to no specialized compiler support.
        Programs are compiled twice; once with Haste and once with GHC, to
        produce the final client and server side code respectively.
\end{itemize}

@@

\begin{code}
clientMain :: App Done
clientMain = do
  greetings <- export helloServer

@@

This is the entry point of our application. Recall that Haste.App applications
are written from the client's perspective, making this the starting point of
the client side code.

@@

Finally, we need to launch our application and inform the client part of the
whereabouts of its server counterpart. In this example the server will listen
for connecting clients on port 1111, and since we will only run this
application locally the client should attempt to contact the server over the
loopback interface.

\begin{code}
main :: IO ()
main = runApp (mkConfig "ws://localhost:1111" 1111)
              clientMain
\end{code}

The URL of the server is provided for the benefit of the client and the port
to listen on is provided for the benefit of the server. The final argument is
the main function of our application. Figure \ref{lst:hello-server} gives the
code of this example in its entirety.

Like \lstinline!runClient!, \lstinline!runApp! does not return, and is intended
to be the only computation executed by \lstinline!main!. In fact, performing
non-deterministic IO and allowing \lstinline!runApp! or its arguments to
capture the result may lead to unpredictable behavior. The \lstinline!clientMain!
function should for all intents and purposes be regarded as the program's entry
point. A non-intrusive method of enforcing this using the type system is
described in section \ref{sec:limitations}, but is not used in the examples
given here as it relies on the use of the GHC plugin system.

@@

clientMain :: App Done
clientMain = do
  greetings <- export helloServer

  runClient $ do
    name <- prompt "Hi there, what is your name?"
    onServer (greetings <.> name)

main :: IO ()
main = runApp (mkConfig "ws://localhost:1111" 1111)
              clientMain
\end{code}
\caption{A seamless programming model: Hello Server}
\label{lst:hello-server}
\end{listingfloat}

@@

While the Hello Server example illustrates how client-server communication is
handled, most web applications need to keep some server side state as well.
At first glance, this would seem to be a trivial problem; can we not just
create our state holding elements in \lstinline!main! as usual?
Figure \ref{lst:bad-state} describes an attempt at a solution using this idea.

\begin{listingfloat}
\begin{code}
main = do
  visitorref <- newIORef 0
  runApp config $ do
    count <- export . liftIO $ do
      atomicModifyIORef visitorref (\v -> (v+1, v+1))

    runClient $ do
      visitors <- onServer count
      alert ("Your are visitor #" ++ show visitors)
\end{code}
\caption{server side state: a misguided first attempt}
\label{lst:bad-state}
\end{listingfloat}

While this may at first seem to solve the problem of server side state, this
approach raises a number of questions. Does \lstinline!visitorref! live on the
client or the server, or perhaps on both? What happens if our state holding
element is initialized in a non-deterministic way, and what if we assign it
a value on the server and then attempt to read that value on the client?
Recall that, as explained in section \ref{sec:helloserver}, the function
passed to \lstinline!runApp! should for all intents and purposes be regarded as
the program's entry point.

To allow the use of server state while treating the final argument to
\lstinline!runApp! (from here on always referred to as \lstinline!clientMain!)
as the program's entry point, we may want to allow arbitrary IO computations to
be lifted into the \lstinline!App! monad:

\begin{code}
main = runApp config $ do
  visitorref <- liftIO $ newIORef 0
  ...
\end{code}

However, this does not answer the question of where \lstinline!visitorref!
lives or where any side effects caused by its creation will take place.
To solve this problem once and for all, we need to introduce a way to lift
arbitrary IO computations, but ensure that said computations are executed on
the server and nowhere else. This is accomplished using a more restricted
version of \lstinline!liftIO!:

\begin{code}
liftServerIO :: IO a -> App (Useless a)
\end{code}

In contrast to \lstinline!liftIO!, the result value of \lstinline!liftServerIO!
is wrapped in the \lstinline!Useless! type. As the name implies, this type is
completely opaque and thus useless unless its contained value can somehow be
extracted, something which is only possible using a single function:

\begin{code}
mkUseful :: Useless a -> Server a
\end{code}

Therein lies the secret to restricting \lstinline!liftServerIO! to the
server: the only function capable of turning a \lstinline!Useless! value into
a useful one lives in the server monad. Any client side code is thus free to
completely ignore executing computations lifted using \lstinline!liftServerIO!;
since the result of a server lifted computation is never observable on the
client, the client has no obligation to even produce such a value.
Figure \ref{lst:good-state} shows how to make proper use of server side state.

\begin{listingfloat}
\begin{code}
main =
  runApp config $ do
    uselessref <- liftServerIO $ newIORef 0

    count <- export $ do
      r <- mkUseful uselessref
      liftIO $ atomicModifyIORef r (\v -> (v+1, v+1))

@@

\begin{code}
srvHello :: Useless RcptList -> Server ()
srvHello uRcpts = do
  recipients <- mkUseful uRcpts
  sid <- getSessionID
  liftIO $ do
    rcptMVar <- CC.newEmptyMVar
    CC.modifyMVar recipients $ \cs ->
      return ((sid, rcptMVar):cs ,())
\end{code}

@@

\begin{code}
srvSend :: Useless RcptList -> String -> Server ()
srvSend uRcpts message = do
    rcpts <- mkUseful uRcpts
    liftIO $ do
      recipients <- CC.readMVar rcpts
      mapM_ (CC.forkIO . deliver message) recipients
  where
    deliver msg (_,rcptMVar) = CC.putMVar rcptMVar msg
\end{code}

@@

\begin{code}
srvAwait :: Useless RcptList -> Server String
srvAwait uRcpts = do
  rcpts <- mkUseful uRcpts
  sid <- getSessionID
  liftIO $ do
    recipients <- CC.readMVar rcpts
    case lookup sid recipients of
      Just mv -> CC.takeMVar mv
      _       -> fail "Unregistered session!"
\end{code}

@@

\begin{code}
clientMain :: App Done
clientMain = do
  recipients <- liftServerIO $ CC.newMVar []

@@

This is the final part of our program; we set up an event handler to clear the
input box and send its contents off to the server whenever the user hits return
(character code 13) while the input box has focus. Finally, all that is left
is to launch our application.

\begin{code}
main :: IO ()
main =
  runApp (mkConfig "ws://localhost:1111" 1111)
         clientMain
\end{code}

The discerning reader may be slightly annoyed at the need to extract the
contents from \lstinline!Useless! values at each point of use. Indeed, in a
simple example such as this, the source clutter caused by this becomes a
disproportionate irritant. Fortunately, most web applications tend to have
more complex client-server interactions, reducing this overhead significantly.

@@

runClient    :: Client () -> App Done
liftServerIO :: IO a -> App (Useless a)
export       :: Exportable a
             => a -> App (Export a)

@@

mkUseful     :: Useless a -> Server a
getSessionID :: Server SessionID
\end{code}
\caption{Types of the Haste.App core functions}
\label{lst:api}
\end{listingfloat}

\begin{table}
\renewcommand{\arraystretch}{1.5}
\begin{center}
\begin{tabular}{|r|l|}
\hline
Function & Purpose \\
\hline
\lstinline!runApp! & Begin execution of a Haste.App program. \\
\lstinline!runClient! & \parbox[t]{5cm}{Lift a single \lstinline!Client!
computation into the \lstinline!App! monad. Must be at the very end of
an \lstinline!App! computation, which is enforced by the type system.} \\
\lstinline!liftServerIO! & \parbox[t]{5cm}{Lift an IO computation into the
\lstinline!App! monad. The computation and its result are exclusive to the
server, as enforced by the type system, and are not observable on the client.} \\
\lstinline!export! & \parbox[t]{5cm}{Make a server side function available to
be called remotely by the client.} \\
\lstinline!onServer! & \parbox[t]{5cm}{Dispatch a remote call to the server and
wait for its completion. The result of the remote computation is returned on
the client after it completes.} \\
\lstinline!<.>! & \parbox[t]{5cm}{Apply an \lstinline!export!ed function to
a serializable argument.} \\
\lstinline!mkUseful! & \parbox[t]{5cm}{Extract the contents of a
\lstinline!Useless! value produced by \lstinline!liftServerIO! on the server.} \\
\lstinline!getSessionID! & \parbox[t]{5cm}{Get the unique identifier for
the current session. This is a pure convenience function, to relieve
programmers of the burden of session bookkeeping.} \\
\hline
\end{tabular}
\end{center}
\caption{Core functions of Haste.App}
\label{tbl:api}
\end{table}

@@

\begin{code}
data Useless a = Useful a | Useless
data Export a = Export CallID [JSON]

liftServerIO :: IO a -> App (Useless a)
export :: Exportable a => a -> App (Export a)
\end{code}

@@

\begin{code}
liftServerIO _ = App $ return Useless
\end{code}

@@

\begin{code}
liftServerIO m = App $ fmap Useful m
\end{code}

Its argument IO computation is executed, and the result is wrapped in the
\lstinline!Useful! constructor. The counterpart function, \lstinline!mkUseful!
simply unwraps the \lstinline!Useful! constructor and returns the contained
value in the \lstinline!Server! monad.

\begin{code}
mkUseful :: Useless a -> Server a
mkUseful (Useful x) = return x
\end{code}

@@

\paragraph{Preventing divergence} While our solution is completely safe from
the invocation point of the \lstinline!runApp! function onward, capturing
IO-dependent variables introduced before this invocation may lead to the client
and server execution paths diverging. Recall that our solution does not modify
either of the compilers used. As such, both will produce code to execute all
computations called by \lstinline!main!. Consider the following example:

\begin{code}
main = do
  exportTwo <- randomRIO (False, True)

  runApp config $ do
    when exportTwo $ do
      _ <- export (return () :: Server ())
      return ()
    fun <- export $ return "I'm a little teapot"

    runClient $ do
      onServer fun >>= alert
\end{code}

Whenever \lstinline!exportTwo! gets assigned a \lstinline!True! value, this
application will call \lstinline!export! twice, assigning the identifier
\lstinline!0! to the first exported computation, which is discarded, and
\lstinline!1! to the second, \lstinline!fun!, which is kept and invoked.
If, on the other hand, \lstinline!exportTwo! is \lstinline!False!,
only the second computation will be exported, and will thus be assigned the
identifier \lstinline!1!.

This is perfectly fine as long as the client and server both assign the
same identifier to \lstinline!fun!. However, since \lstinline!randomRIO! is
not deterministic and since all computations called by \lstinline!main! will be
executed on both the client and the server, there is a possibility that
\lstinline!fun! will be assigned the identifier \lstinline!0! on the client and
\lstinline!1! on the server. In this scenario, when the client attempts to call
the \lstinline!return "I'm a little teapot"! computation over the network,
the server instead executes the \lstinline!return ()! computation, which
returns \lstinline!()! rather than \lstinline!String!, breaking the type safety
of our program!

This is easily remedied however, by a slight compiler modification. Instead of
using \lstinline!main :: IO ()! as the program entry point, we introduce a
function \lstinline!clientMain :: App Done! for this purpose, and let the runtime
system start the application by invoking \lstinline!runApp config clientMain!,
where \lstinline!config! contains the URL and port configuration for the
server, given by the programmer using a compiler flag.

As this modification only affects the program's entry, point which is already
given ``special'' treatment in Haskell compilers, and may even be implemented
using the plugin mechanism of GHC, requiring no modification of the compiler
itself, we believe this to be a practical, non-intrusive way to remove this
limitation.

The astute reader may notice that the same applies to code using
\lstinline!unsafePerformIO! to declare IO-dependent top-level values, and that
this problem is unaffected by changing the type of the program's entry point.
While this is indeed true, we feel that since it is quite possible to break
any Haskell program using \lstinline!unsafePerformIO!, even subverting the type
system, ``solving'' \lstinline!unsafePerformIO! is not within the scope of this
work.

>>>>

\begin{itemize}
  \item The programming model is synchronous, giving the programmer a simple,
        linear view of the program flow, eliminating the need to program with
        callbacks and continuations.
  \item Side-effecting code is explicitly designated to run on either the
        client or the server using the type system while pure code can be
        shared by both. Additionally, general IO computations may be lifted
        into both client and server code, allowing for safe IO code reuse
        within the confines of the client or server designated functions.
  \item Client-server network communication is handled through statically typed
        RPC function calls, extending the reach of Haskell's type checker over
        the network and giving the programmer advance warning when she uses
        network services incorrectly or forgets to update communication code
        as the application's internal protocol changes.
  \item Our model takes the view that the client side is the main driver when
        developing web applications and accordingly assigns the server
        the role of a computational and/or storage resource, tasked with
        servicing client requests rather than driving the program. While it is
        entirely possible to implement a server-to-client communication channel
        on top of our model, we believe that choosing one side of the
        heterogenous client-server relation as the master helps keeping the
        program flow linear and predictable.
  \item The implementation is built as a library on top of the GHC and Haste
        Haskell compilers, requiring little to no specialized compiler support.
        Programs are compiled twice; once with Haste and once with GHC, to
        produce the final client and server side code respectively.
\end{itemize}

@@

\begin{code}
main :: App Done
main = do
  greetings <- export helloServer

@@

The \lstinline!main! function is, as usual, the entry point of our application.
In contrast to traditional applications which live either on the client or on
the server and begin in the \lstinline!IO! monad, Haste.App applications live
on both and begin execution in the \lstinline!App! monad which provides some
crucial tools to facilitate typed communication between the two.

@@

Figure \ref{lst:hello-server} gives the code of this example in its entirety.

@@

main :: App Done
main = do
  greetings <- export helloServer

  runClient $ do
    name <- prompt "Hi there, what is your name?"
    onServer (greetings <.> name)
\end{code}
\caption{A seamless programming model: Hello Server}
\label{lst:hello-server}
\end{listingfloat}

@@

While the Hello Server example illustrates how client-server communication is
handled, most web applications need to keep some server side state as well.
How can we create state holding elements for the server which are not
accessible to the client?

To accomplish this, we need to introduce a way to lift arbitrary IO
computations, but ensure that said computations are executed on the server and
nowhere else. This is accomplished using a more restricted version of
\lstinline!liftIO!:

\begin{code}
liftServerIO :: IO a -> App (Remote a)
\end{code}

In contrast to \lstinline!liftIO!, the result value of \lstinline!liftServerIO!
is wrapped in the abstract \lstinline!Remote! type. As the name implies,
values of this type live exclusively on the server. In order to inspect such a
value is needs to be extracted somehow, something which is only possible using
a single function:

\begin{code}
fromRemote :: Remote a -> Server a
\end{code}

Therein lies the secret to restricting \lstinline!liftServerIO! to the
server: the only function capable of turning a \lstinline!Remote! value into
a useful one lives in the server monad. Any client side code is thus free to
completely ignore executing computations lifted using \lstinline!liftServerIO!;
since the result of a server lifted computation is never observable on the
client, the client has no obligation to even produce such a value.
Figure \ref{lst:good-state} shows how to make proper use of server side state.

\begin{listingfloat}
\begin{code}
main = do
  remoteref <- liftServerIO $ newIORef 0

  count <- export $ do
    r <- fromRemote remoteref
    liftIO $ atomicModifyIORef r (\v -> (v+1, v+1))

@@

\begin{code}
srvHello :: Remote RcptList -> Server ()
srvHello remoteRcpts = do
  recipients <- fromRemote remoteRcpts
  sid <- getSessionID
  liftIO $ do
    rcptMVar <- CC.newEmptyMVar
    CC.modifyMVar recipients $ \cs ->
      return ((sid, rcptMVar):cs ,())
\end{code}

@@

\begin{code}
srvSend :: Remote RcptList -> String -> Server ()
srvSend remoteRcpts message = do
    rcpts <- fromRemote remoteRcpts
    liftIO $ do
      recipients <- CC.readMVar rcpts
      mapM_ (CC.forkIO . deliver message) recipients
  where
    deliver msg (_,rcptMVar) = CC.putMVar rcptMVar msg
\end{code}

@@

\begin{code}
srvAwait :: Remote RcptList -> Server String
srvAwait remoteRcpts = do
  rcpts <- fromRemote remoteRcpts
  sid <- getSessionID
  liftIO $ do
    recipients <- CC.readMVar rcpts
    case lookup sid recipients of
      Just mv -> CC.takeMVar mv
      _       -> fail "Unregistered session!"
\end{code}

@@

\begin{code}
main :: App Done
main = do
  recipients <- liftServerIO $ CC.newMVar []

@@

This is the final part of our program; we set up an event handler to clear the
input box and send its contents off to the server whenever the user hits return
(character code 13) while the input box has focus.

The discerning reader may be slightly annoyed at the need to extract the
contents from \lstinline!Remote! values at each point of use. Indeed, in a
simple example such as this, the source clutter caused by this becomes a
disproportionate irritant. Fortunately, most web applications tend to have
more complex client-server interactions, reducing this overhead significantly.

@@

runClient    :: Client () -> App Done
liftServerIO :: IO a -> App (Remote a)
export       :: Exportable a
             => a -> App (Export a)

@@

fromRemote   :: Remote a -> Server a
getSessionID :: Server SessionID
\end{code}
\caption{Types of the Haste.App core functions}
\label{lst:api}
\end{listingfloat}

\begin{table}
\renewcommand{\arraystretch}{1.5}
\begin{center}
\begin{tabular}{|r|l|}
\hline
Function & Purpose \\
\hline
\lstinline!runApp! & Begin execution of a Haste.App program. \\
\lstinline!runClient! & \parbox[t]{5cm}{Lift a single \lstinline!Client!
computation into the \lstinline!App! monad. Must be at the very end of
an \lstinline!App! computation, which is enforced by the type system.} \\
\lstinline!liftServerIO! & \parbox[t]{5cm}{Lift an IO computation into the
\lstinline!App! monad. The computation and its result are exclusive to the
server, as enforced by the type system, and are not observable on the client.} \\
\lstinline!export! & \parbox[t]{5cm}{Make a server side function available to
be called remotely by the client.} \\
\lstinline!onServer! & \parbox[t]{5cm}{Dispatch a remote call to the server and
wait for its completion. The result of the remote computation is returned on
the client after it completes.} \\
\lstinline!<.>! & \parbox[t]{5cm}{Apply an \lstinline!export!ed function to
a serializable argument.} \\
\lstinline!fromRemote! & \parbox[t]{5cm}{Extract the contents of a
\lstinline!Remote! value produced by \lstinline!liftServerIO! on the server.} \\
\lstinline!getSessionID! & \parbox[t]{5cm}{Get the unique identifier for
the current session. This is a pure convenience function, to relieve
programmers of the burden of session bookkeeping.} \\
\hline
\end{tabular}
\end{center}
\caption{Core functions of Haste.App}
\label{tbl:api}
\end{table}

@@

\begin{code}
data Remote a = Here a | There
data Export a = Export CallID [JSON]

liftServerIO :: IO a -> App (Remote a)
export :: Exportable a => a -> App (Export a)
\end{code}

@@

\begin{code}
liftServerIO _ = App $ return There
\end{code}

@@

\begin{code}
liftServerIO m = App $ fmap Here m
\end{code}

Its argument IO computation is executed, and the result is wrapped in the
\lstinline!Remote! constructor. The counterpart function, \lstinline!fromRemote!
simply unwraps the \lstinline!Here! constructor and returns the contained
value in the \lstinline!Server! monad.

\begin{code}
fromRemote :: Remote a -> Server a
fromRemote (Here x) = return x
\end{code}

@@
